---
title: "Network Analysis in Tairona Chiefdoms of the Río Frío Basin"
subtitle: "Datasets by ceramic typology and by mineral temper for the three occupation periods"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    df_print: paged
---
# 1. Packages used in R language for network analysis
```{r message=TRUE, warning=TRUE}
# (A) Load required packages
required_packages <- c(
  "sf", "spdep", "ggplot2", "viridis", "car",
  "spatstat.explore", "spatstat.geom", "spatstat.model", 
  "network", "ggspatial", "vegan", "scales", "dplyr", 
  "patchwork", "tidygraph", "ggraph", "GGally", "FSA",
  "stats", "ggpubr", "writexl", "openxlsx", "sna", 
  "statnet", "igraph", "DT", "tnet", "kableExtra",
  "readxl", "dunn.test", "ggdist"
)

# Install missing packages if necessary
for (pkg in required_packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg)
  }
}

# Load all required libraries
invisible(lapply(required_packages, library, character.only = TRUE))

cat("All packages have been successfully installed and loaded.\n\n")
```

# 2. Definition of the working environment
```{r message=FALSE, warning=FALSE}
# List of Excel files in the repository
xlsx_files <- c(
  "Buritaca_Desgrasantes.xlsx",
  "Buritaca_Desgrasantes_attr.xlsx",
  "Buritaca_Tipos.xlsx",
  "Buritaca_Tipos_attr.xlsx",
  "Neguanje_Desgrasantes.xlsx",
  "Neguanje_Desgrasantes_attr.xlsx",
  "Neguanje_Tipos.xlsx",
  "Neguanje_Tipos_attr.xlsx",
  "Tairona_Desgrasantes.xlsx",
  "Tairona_Desgrasantes_attr.xlsx",
  "Tairona_Tipos.xlsx",
  "Tairona_Tipos_attr.xlsx"
)

github_xlsx_base <- "https://raw.githubusercontent.com/luismiguelsoto/Network_Analysis_in_Tairona_Chiefdoms/main/Database/"

read_xlsx_from_github <- function(filename) {
  url <- paste0(github_xlsx_base, utils::URLencode(filename, reserved = TRUE))
  cat("Downloading from:", url, "\n")
  tmp <- tempfile(fileext = ".xlsx")
  download.file(url, destfile = tmp, mode = "wb")
  df <- read_excel(tmp)
  unlink(tmp)
  return(df)
}

data_list <- lapply(xlsx_files, read_xlsx_from_github)
names(data_list) <- sub("\\.xlsx$", "", xlsx_files)
cat("\nAll Excel files have been downloaded and read successfully.\n")

# For convenience, create shorter variable names
Neguanje_Tipos             <- data_list[["Neguanje_Tipos"]]
Neguanje_Tipos_attr        <- data_list[["Neguanje_Tipos_attr"]]
Buritaca_Tipos             <- data_list[["Buritaca_Tipos"]]
Buritaca_Tipos_attr        <- data_list[["Buritaca_Tipos_attr"]]
Tairona_Tipos              <- data_list[["Tairona_Tipos"]]
Tairona_Tipos_attr         <- data_list[["Tairona_Tipos_attr"]]
Neguanje_Desgrasantes      <- data_list[["Neguanje_Desgrasantes"]]
Neguanje_Desgrasantes_attr <- data_list[["Neguanje_Desgrasantes_attr"]]
Buritaca_Desgrasantes      <- data_list[["Buritaca_Desgrasantes"]]
Buritaca_Desgrasantes_attr <- data_list[["Buritaca_Desgrasantes_attr"]]
Tairona_Desgrasantes       <- data_list[["Tairona_Desgrasantes"]]
Tairona_Desgrasantes_attr  <- data_list[["Tairona_Desgrasantes_attr"]]

# Group datasets for threshold analysis (Figure 2)
datasets <- list(
  Neguanje_Tipos = Neguanje_Tipos,
  Neguanje_Tipos_attr = Neguanje_Tipos_attr,
  Buritaca_Tipos = Buritaca_Tipos,
  Buritaca_Tipos_attr = Buritaca_Tipos_attr,
  Tairona_Tipos = Tairona_Tipos,
  Tairona_Tipos_attr = Tairona_Tipos_attr,
  Neguanje_Desgrasantes = Neguanje_Desgrasantes,
  Neguanje_Desgrasantes_attr = Neguanje_Desgrasantes_attr,
  Buritaca_Desgrasantes = Buritaca_Desgrasantes,
  Buritaca_Desgrasantes_attr = Buritaca_Desgrasantes_attr,
  Tairona_Desgrasantes = Tairona_Desgrasantes,
  Tairona_Desgrasantes_attr = Tairona_Desgrasantes_attr
)
```

# 3. Creation of figures in descending order (Network analysis by occupation period)
## 3.1. Figure 3
```{r message=FALSE, warning=FALSE}
# Functions to compute similarity matrix and betweenness centrality based on Peeples' methodology

# References:
# 1. Peeples, M. A. (2017). Network science and statistical techniques for dealing with uncertainties in archaeological datasets.
# 2. Peeples, M. A. (2019). A brief introduction to archaeological networks in R.

# Adjust labels in the attribute data
normalize_site_types <- function(attr_data) {
  attr_data$Site_Type <- trimws(tolower(attr_data$Site_Type))
  attr_data$Site_Type <- ifelse(attr_data$Site_Type == "large sites", "Large sites",
                                ifelse(attr_data$Site_Type == "medium sites", "Medium sites",
                                       "Small sites"))
  return(attr_data)
}

Neguanje_Tipos_attr        <- normalize_site_types(Neguanje_Tipos_attr)
Buritaca_Tipos_attr        <- normalize_site_types(Buritaca_Tipos_attr)
Tairona_Tipos_attr         <- normalize_site_types(Tairona_Tipos_attr)
Neguanje_Desgrasantes_attr <- normalize_site_types(Neguanje_Desgrasantes_attr)
Buritaca_Desgrasantes_attr <- normalize_site_types(Buritaca_Desgrasantes_attr)
Tairona_Desgrasantes_attr  <- normalize_site_types(Tairona_Desgrasantes_attr)

# Define function to calculate Brainerd-Robinson similarity matrix
sim.mat <- function(x) {
  names <- row.names(x)
  x <- na.omit(x)
  x <- prop.table(as.matrix(x), 1)
  rd <- dim(x)[1]
  results <- matrix(0, rd, rd)
  for (s1 in 1:rd) {
    for (s2 in 1:rd) {
      x1Temp <- as.numeric(x[s1, ])
      x2Temp <- as.numeric(x[s2, ])
      results[s1, s2] <- 2 - (sum(abs(x1Temp - x2Temp)))
    }
  }
  row.names(results) <- names
  colnames(results) <- names
  results <- round(results / 2, 3)
  return(results)
}

# Function to calculate Mantel test
mantel_test <- function(matrix1, matrix2, permutations = 10000) {
  result <- vegan::mantel(matrix1, matrix2, permutations = permutations)
  return(data.frame(
    Mantel_R = result$statistic,
    P_Value = result$signif
  ))
}

# Define the range of thresholds to test
thresholds <- seq(0.5, 0.9, by = 0.05)

# Initialize a data frame to store Mantel test results
mantel_results <- data.frame()

# Iterate over each dataset and calculate Mantel test between thresholds
for (dataset_name in names(datasets)) {
  cat("\nProcessing Mantel test for:", dataset_name, "\n")
  
  data <- datasets[[dataset_name]]
  matrices <- list()
  
  # Generate similarity matrices for each threshold
  for (thresh in thresholds) {
    sim_matrix <- tryCatch({
      event2dichot(sim.mat(data), method = "absolute", thresh = thresh)
    }, error = function(e) NULL)
    
    if (!is.null(sim_matrix)) matrices[[as.character(thresh)]] <- sim_matrix
  }
  
  # Compare consecutive matrices using Mantel test
  if (length(matrices) > 1) {
    for (i in 1:(length(matrices) - 1)) {
      matrix1 <- as.dist(matrices[[i]])
      matrix2 <- as.dist(matrices[[i + 1]])
      mantel_result <- mantel_test(matrix1, matrix2)
      mantel_results <- rbind(mantel_results, data.frame(
        Dataset = dataset_name,
        Threshold_1 = names(matrices)[i],
        Threshold_2 = names(matrices)[i + 1],
        Mantel_R = mantel_result$Mantel_R,
        P_Value = mantel_result$P_Value
      ))
    }
  }
}

# Check the structure of mantel_results
str(mantel_results)

# Ensure Threshold_1 and Mantel_R are numeric
mantel_results$Threshold_1 <- as.numeric(as.character(mantel_results$Threshold_1))
mantel_results$Mantel_R <- as.numeric(as.character(mantel_results$Mantel_R))

# Remove missing or invalid data if any
if (any(is.na(mantel_results$Threshold_1))) {
  mantel_results <- mantel_results[!is.na(mantel_results$Threshold_1), ]
}

# Plot Mantel test results
ggplot(mantel_results, aes(x = as.numeric(Threshold_1), y = Mantel_R, group = Dataset, color = Dataset)) +
  geom_line(size = 1) +
  geom_point(size = 3, shape = 21, fill = "white", stroke = 1) +
  scale_color_viridis_d(option = "plasma", guide = guide_legend(ncol = 1)) +
  scale_x_continuous(breaks = pretty_breaks(n = 10), labels = scales::comma) +
  scale_y_continuous(breaks = pretty_breaks(n = 10)) +
  theme_classic(base_size = 14) +
  theme(
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size = 10),
    legend.key.width = unit(1, "cm"),
    legend.key.height = unit(0.5, "cm"),
    legend.spacing.x = unit(0.5, "cm"),
    legend.spacing.y = unit(0.2, "cm"),
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
    axis.title.x = element_text(face = "bold", size = 14),
    axis.title.y = element_text(face = "bold", size = 14),
    panel.grid.major = element_line(color = "gray90")
  ) +
  labs(
    title = "Mantel Coefficient between Similarity Thresholds",
    subtitle = "Relationship between similarity matrices at consecutive thresholds",
    x = "Initial Threshold",
    y = "Mantel Coefficient (R)",
    color = "Datasets by Occupation Period"
  )

# Render interactive table
datatable(
  mantel_results,
  extensions = 'Buttons',
  options = list(dom = 'Bfrtip', buttons = c('csv', 'excel')),
  rownames = FALSE
)
```

## 3.1. Figure 4 and Figure 5 (functions definition)
```{r}
# Helper function to retrieve period duration
get_duration <- function(period_name) {
  durations <- data.frame(
    Period   = c("Neguanje", "Buritaca", "Tairona"),
    Duration = c(700, 400, 700)  # Duration in years / for Neguanje 700; for Buritaca 400; for Tairona 700
  )
  durations$Duration[durations$Period == period_name]
}

# Main function: calculates betweenness, normalizes, transforms, plots, and returns a table
plot_betweenness_scaled <- function(
  data,              # e.g. Neguanje_Tipos
  attr_data,         # e.g. Neguanje_Tipos_attr
  period,            # "Neguanje", "Buritaca", or "Tairona"
  attribute_type,    # "Tipos" or "Tempers"
  threshold   = 0.75,
  size_factor = 5
) {
  # A) Build Brainerd-Robinson matrix and network
  BR_matrix <- sim.mat(data)
  BR_net    <- network(event2dichot(BR_matrix, method = "absolute", thresh = threshold), directed = FALSE)
  BR_net %v% "vertex.names" <- row.names(BR_matrix)
  
  # B) Calculate betweenness and normalize by period duration
  bw <- sna::betweenness(BR_net, gmode = "graph")
  period_duration <- get_duration(period)
  normalized_bw   <- bw / period_duration
  
  # C) Optional sqrt transform to avoid extremes
  transformed_bw  <- sqrt(normalized_bw)
  
  # D) Color by Site_Type
  reg.col <- ifelse(attr_data$Site_Type == "Large sites", "red",
                    ifelse(attr_data$Site_Type == "Medium sites", "green", "blue"))
  
  # E) Edge weights for visualization
  BR_matrix_logged <- log(BR_matrix + 1)
  
  # F) Node size by (transformed) betweenness
  vertex_size <- transformed_bw * size_factor
  
  # G) Plot the network (base R)
  plot_title <- paste0(period, " ", attribute_type, " | Normalized Betweenness")
  plot(BR_net,
       edge.col      = adjustcolor("gray70", alpha.f = 0.5),
       edge.lwd      = BR_matrix_logged,
       vertex.col    = reg.col,
       vertex.border = "black",
       vertex.cex    = vertex_size,
       displaylabels = FALSE,
       main          = plot_title)
  
  legend("topright",
         legend = c("Large Sites", "Medium Sites", "Small Sites"),
         col    = c("red", "green", "blue"),
         pch    = 16,
         pt.cex = c(1.5, 1.2, 0.9),
         bty    = "n",
         title  = "Site Types")
  
  # H) Return a table with site-level betweenness
  site_names <- row.names(BR_matrix)
  site_types <- attr_data$Site_Type
  
  df <- data.frame(
    Site            = site_names,
    Site_Type       = site_types,
    Betweenness     = round(bw, 3),
    Normalized_Betw = round(normalized_bw, 6),
    Period          = period,
    Attribute_Type  = attribute_type,
    stringsAsFactors = FALSE
  )
  
  return(df)
}
```

## 3.2. Figure 4 (network plots: ceramic types)
```{r}
par(mfrow = c(1, 3))

df_Neguanje_Tipos <- plot_betweenness_scaled(
  data          = Neguanje_Tipos,
  attr_data     = Neguanje_Tipos_attr,
  period        = "Neguanje",
  attribute_type= "Tipos",
  threshold     = 0.75,
  size_factor   = 5
)

df_Buritaca_Tipos <- plot_betweenness_scaled(
  data          = Buritaca_Tipos,
  attr_data     = Buritaca_Tipos_attr,
  period        = "Buritaca",
  attribute_type= "Tipos",
  threshold     = 0.75,
  size_factor   = 5
)

df_Tairona_Tipos <- plot_betweenness_scaled(
  data          = Tairona_Tipos,
  attr_data     = Tairona_Tipos_attr,
  period        = "Tairona",
  attribute_type= "Tipos",
  threshold     = 0.75,
  size_factor   = 5
)

par(mfrow = c(1, 1))
```

## 3.2.1. Figure 4 (normalized data set for ceramic types by site typology)
```{r}
# Combine the three data frames for Neguanje, Buritaca, Tairona (ceramic types)
betweenness_table_tipos <- do.call(rbind, list(
  df_Neguanje_Tipos,
  df_Buritaca_Tipos,
  df_Tairona_Tipos
))

# Display the combined table with DT
DT::datatable(
  betweenness_table_tipos,
  extensions = 'Buttons',
  options = list(dom = 'Bfrtip', buttons = c('csv', 'excel')),
  rownames = FALSE
)

# Reorder the factor "Period" for the desired order
betweenness_table_tipos$Period <- factor(
  betweenness_table_tipos$Period, 
  levels = c("Neguanje", "Buritaca", "Tairona")
)
```

## 3.2.2. Figure 4 ()
```{r}
ggplot(betweenness_table_tipos, aes(x = Site_Type, y = Normalized_Betw, fill = Site_Type)) +
  geom_boxplot(width = 0.5, outlier.shape = NA, alpha = 0.7, color = "black") +
  geom_jitter(width = 0.1, alpha = 0.5, shape = 21, color = "black") +
  facet_wrap(~ Period, ncol = 1) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Normalized Betweenness by Site Type",
       x = "Site Type", y = "Normalized Betweenness") +
  coord_flip() +
  theme_bw(base_size = 14) +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold", size = 14)
  )
```

## 3.3. Figure 5 (network plots: ceramic tempers)
```{r}
par(mfrow = c(1, 3))

df_Neguanje_Tempers <- plot_betweenness_scaled(
  data          = Neguanje_Desgrasantes,
  attr_data     = Neguanje_Desgrasantes_attr,
  period        = "Neguanje",
  attribute_type= "Tempers",
  threshold     = 0.75,
  size_factor   = 5
)

df_Buritaca_Tempers <- plot_betweenness_scaled(
  data          = Buritaca_Desgrasantes,
  attr_data     = Buritaca_Desgrasantes_attr,
  period        = "Buritaca",
  attribute_type= "Tempers",
  threshold     = 0.75,
  size_factor   = 5
)

df_Tairona_Tempers <- plot_betweenness_scaled(
  data          = Tairona_Desgrasantes,
  attr_data     = Tairona_Desgrasantes_attr,
  period        = "Tairona",
  attribute_type= "Tempers",
  threshold     = 0.75,
  size_factor   = 5
)

par(mfrow = c(1, 1))

betweenness_table_tempers <- do.call(rbind, list(
  df_Neguanje_Tempers,
  df_Buritaca_Tempers,
  df_Tairona_Tempers
))
```

## 3.3.1. Figure 5 (normalized data set for mineral temper by site typology)
```{r}
# Combine the three data frames for Neguanje, Buritaca, Tairona (ceramic types)
betweenness_table_tempers <- do.call(rbind, list(
  df_Neguanje_Tempers,
  df_Buritaca_Tempers,
  df_Tairona_Tempers
))

# Display the combined table with DT
DT::datatable(
  betweenness_table_tempers,
  extensions = 'Buttons',
  options = list(dom = 'Bfrtip', buttons = c('csv', 'excel')),
  rownames = FALSE
)

# Reorder the factor "Period" for the desired order
betweenness_table_tempers$Period <- factor(
  betweenness_table_tempers$Period, 
  levels = c("Neguanje", "Buritaca", "Tairona")
)
```

## 3.3.2. Figure 5 (Raincloud Plot)
```{r}
ggplot(betweenness_table_tempers, aes(x = Site_Type, y = Normalized_Betw, fill = Site_Type)) +
  geom_boxplot(width = 0.5, outlier.shape = NA, alpha = 0.7, color = "black") +
  geom_jitter(width = 0.1, alpha = 0.5, shape = 21, color = "black") +
  facet_wrap(~ Period, ncol = 1) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Normalized Betweenness by Site Type - Tempers",
    x = "Site Type", 
    y = "Normalized Betweenness"
  ) +
  coord_flip() +
  theme_bw(base_size = 14) +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold", size = 14)
  )
```

## 3.5. Figure 6 (normalization by centuries and definition of functions)
```{r}
# Durations by period
durations <- data.frame(
  Period = c("Neguanje", "Buritaca", "Tairona"),
  Duration = c(700, 400, 700) # Duration in years / for Neguanje 700; for Buritaca 400; for Tairona 700
)

# Function to calculate normalized betweenness centrality and create plots
analyze_and_plot_centrality_normalized <- function(data, attr_data, title, period_duration, layout_type = "stress") {
  BR_matrix <- sim.mat(data)
  BR_net <- network(event2dichot(BR_matrix, method = "absolute", thresh = 0.75), directed = FALSE)
  BR_net %v% "vertex.names" <- row.names(BR_matrix)
  bw <- sna::betweenness(BR_net, gmode = "graph")
  
  # Normalize betweenness by period duration
  normalized_bw <- bw / period_duration
  
  BR_net_tidy <- as_tbl_graph(BR_net) %>%
    mutate(betweenness = normalized_bw) %>%
    mutate(size = betweenness, color = betweenness)
  
  p <- ggraph(BR_net_tidy, layout = layout_type) +
    geom_edge_link(alpha = 0.5, color = "gray70") +
    geom_node_point(aes(size = size, color = color)) +
    scale_color_viridis_c(name = "Normalized\nBetweenness") +
    theme_graph() +
    ggtitle(title)
  
  return(p)
}
```

## 3.6.1. Figure 4 (Neguanje period)
```{r message=FALSE, warning=FALSE}
# Ceramic Types
analyze_and_plot_centrality_normalized(
  data = Neguanje_Tipos,
  attr_data = Neguanje_Tipos_attr,
  title = "Neguanje Period | Ceramic Types",
  period_duration = durations$Duration[durations$Period == "Neguanje"]
)

# Tempers
analyze_and_plot_centrality_normalized(
  data = Neguanje_Desgrasantes,
  attr_data = Neguanje_Desgrasantes_attr,
  title = "Neguanje Period | Tempers",
  period_duration = durations$Duration[durations$Period == "Neguanje"]
)
```

## 3.6.2. Figure 4 (Buritaca period)
```{r message=FALSE, warning=FALSE}
# Ceramic Types
analyze_and_plot_centrality_normalized(
  data = Buritaca_Tipos,
  attr_data = Buritaca_Tipos_attr,
  title = "Buritaca Period | Ceramic Types",
  period_duration = durations$Duration[durations$Period == "Buritaca"]
)

# Tempers
analyze_and_plot_centrality_normalized(
  data = Buritaca_Desgrasantes,
  attr_data = Buritaca_Desgrasantes_attr,
  title = "Buritaca Period | Tempers",
  period_duration = durations$Duration[durations$Period == "Buritaca"]
)
```

## 3.6.3. Figure 4 (Tairona period)
```{r message=FALSE, warning=FALSE}
# Ceramic Types
analyze_and_plot_centrality_normalized(
  data = Tairona_Tipos,
  attr_data = Tairona_Tipos_attr,
  title = "Tairona Period | Ceramic Types",
  period_duration = durations$Duration[durations$Period == "Tairona"]
)

# Tempers
analyze_and_plot_centrality_normalized(
  data = Tairona_Desgrasantes,
  attr_data = Tairona_Desgrasantes_attr,
  title = "Tairona Period | Tempers",
  period_duration = durations$Duration[durations$Period == "Tairona"]
)
```

## 3.7 Figure 5 (function definitions and normalized calculations by period)
```{r}
# Function to calculate betweenness centrality
calc_betweenness <- function(data) {
  BR_matrix <- sim.mat(data) # Calculate Brainerd-Robinson similarity matrix
  BR_net <- network(event2dichot(BR_matrix, method = "absolute", thresh = 0.75), directed = FALSE)
  return(sna::betweenness(BR_net, gmode = "graph")) # Calculate betweenness centrality
}

# Calculate betweenness centrality
Neguanje_BC <- calc_betweenness(Neguanje_Tipos)
Buritaca_BC <- calc_betweenness(Buritaca_Tipos)
Tairona_BC <- calc_betweenness(Tairona_Tipos)
Neguanje_Temp_BC <- calc_betweenness(Neguanje_Desgrasantes)
Buritaca_Temp_BC <- calc_betweenness(Buritaca_Desgrasantes)
Tairona_Temp_BC <- calc_betweenness(Tairona_Desgrasantes)

# Combine into a single data frame
combined_data <- rbind(
  data.frame(Betweenness = Neguanje_BC, Attribute = "Ceramic_Types", Period = "Neguanje"),
  data.frame(Betweenness = Buritaca_BC, Attribute = "Ceramic_Types", Period = "Buritaca"),
  data.frame(Betweenness = Tairona_BC, Attribute = "Ceramic_Types", Period = "Tairona"),
  data.frame(Betweenness = Neguanje_Temp_BC, Attribute = "Tempers", Period = "Neguanje"),
  data.frame(Betweenness = Buritaca_Temp_BC, Attribute = "Tempers", Period = "Buritaca"),
  data.frame(Betweenness = Tairona_Temp_BC, Attribute = "Tempers", Period = "Tairona")
)

# Normalize by duration
combined_data <- combined_data %>%
  left_join(durations, by = "Period") %>%
  mutate(Normalized_Betweenness = Betweenness / Duration)

# Verify the structure of the combined assembly after normalization
str(combined_data)
```

## 3.7.1. Figure 5 (Histogram: ceramic types)
```{r}
# Custom colors and period labels
custom_colors <- c("Neguanje" = "#1f78b4", "Buritaca" = "#33a02c", "Tairona" = "#ff7f00")
period_labels <- c(
  "Neguanje" = "Neguanje (1st-7th Century A.D.)",
  "Buritaca" = "Buritaca (7th-10th Century A.D.)",
  "Tairona" = "Tairona (10th-16th Century A.D.)"
)

# Histogram for Ceramic Types
hist_ceramic_facet <- ggplot(
  combined_data %>% filter(Attribute == "Ceramic_Types") %>%
    mutate(Period = factor(Period, levels = c("Neguanje", "Buritaca", "Tairona"))),
  aes(x = Normalized_Betweenness, fill = Period)
) +
  geom_histogram(
    bins = 30,
    alpha = 0.7,
    color = "black"
  ) +
  scale_fill_manual(values = custom_colors) +
  labs(
    title = "Distribution of Normalized Betweenness Centrality (Ceramic Types)",
    x = "Normalized Betweenness Centrality",
    y = "Frequency"
  ) +
  facet_wrap(~ Period, scales = "free_y", labeller = labeller(Period = period_labels)) +
  theme_bw(base_size = 10) +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 10), # Smaller legend title
    legend.text = element_text(size = 8),  # Smaller legend text
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

# Show the plot
print(hist_ceramic_facet)
```

## 3.7.1. Figure 5 (Histogram: mineral tempers)
```{r}
# Histogram for Tempers
hist_tempers_facet <- ggplot(
  combined_data %>% filter(Attribute == "Tempers") %>%
    mutate(Period = factor(Period, levels = c("Neguanje", "Buritaca", "Tairona"))),
  aes(x = Normalized_Betweenness, fill = Period)
) +
  geom_histogram(
    bins = 30,
    alpha = 0.7,
    color = "black"
  ) +
  scale_fill_manual(values = custom_colors) +
  labs(
    title = "Distribution of Normalized Betweenness Centrality (Tempers)",
    x = "Normalized Betweenness Centrality",
    y = "Frequency"
  ) +
  facet_wrap(~ Period, scales = "free_y", labeller = labeller(Period = period_labels)) +
  theme_bw(base_size = 10) +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 10), # Smaller legend title
    legend.text = element_text(size = 8),  # Smaller legend text
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

# Show the plot
print(hist_tempers_facet)
```

## 3.8. Figure 6
```{r message=FALSE, paged.print=FALSE}
# Define custom colors
custom_colors <- c("Neguanje" = "#1f78b4", "Buritaca" = "#33a02c", "Tairona" = "#ff7f00")

# Reorder Period factor to ensure correct order in the plots
combined_data <- combined_data %>%
  mutate(Period = factor(Period, levels = c("Neguanje", "Buritaca", "Tairona")))

# Boxplot for Ceramic Types
boxplot_ceramic <- ggplot(
  combined_data %>% filter(Attribute == "Ceramic_Types"),
  aes(x = Period, y = Normalized_Betweenness, fill = Period)
) +
  geom_boxplot(
    notch = FALSE,
    outlier.shape = 21,
    outlier.fill = "red",
    alpha = 0.8,
    width = 0.2  # Adjust this value based on the desired width
  ) +
  scale_fill_manual(values = custom_colors) +
  labs(
    title = "Normalized Betweenness 
    Centrality (Ceramic Types)",
    x = "Period of Occupation",
    y = "Normalized Betweenness"
  ) +
  theme_light(base_size = 10)

# Boxplot for Tempers
boxplot_tempers <- ggplot(
  combined_data %>% filter(Attribute == "Tempers"),
  aes(x = Period, y = Normalized_Betweenness, fill = Period)
) +
  geom_boxplot(
    notch = FALSE,
    outlier.shape = 22,
    outlier.fill = "red",
    alpha = 0.8,
    width = 0.2
  ) +
  scale_fill_manual(values = custom_colors) +
  labs(
    title = "Normalized Betweenness 
    Centrality (Tempers)",
    x = "Period of Occupation",
    y = "Normalized Betweenness"
  ) +
  theme_light(base_size = 10)

# Combine and display the plots
combined_plot <- boxplot_ceramic + boxplot_tempers + plot_layout(ncol = 2)
print(combined_plot)

```

## 3.9.1. Non-parametric tests for the normalized data set by occupation period (ceramic types)
```{r}
# Perform Kruskal-Wallis test for Ceramic Types
kruskal_ceramic <- combined_data %>%
  filter(Attribute == "Ceramic_Types") %>%
  na.omit() %>%
  kruskal.test(Normalized_Betweenness ~ Period, data = .)

# Print Kruskal-Wallis test results
cat("**Kruskal-Wallis Test Results for Ceramic Types:**\n")
cat(sprintf("Chi-squared = %.3f, df = %d, p-value = %.5f\n\n",
            kruskal_ceramic$statistic, kruskal_ceramic$parameter, kruskal_ceramic$p.value))

# Check if the test is significant
if (kruskal_ceramic$p.value < 0.05) {
  cat("The Kruskal-Wallis test is significant (p < 0.05). Performing Dunn post hoc test...\n\n")
  
  # Perform Dunn post hoc test
  dunn_ceramic <- combined_data %>%
    filter(Attribute == "Ceramic_Types") %>%
    na.omit() %>%
    dunnTest(Normalized_Betweenness ~ Period, data = ., method = "bonferroni")
  
  # Convert results to data frame for display
  dunn_ceramic_df <- as.data.frame(dunn_ceramic$res)
  
  # Print the Dunn test results in R Markdown
  cat("**Dunn Test Results for Ceramic Types:**\n")
  knitr::kable(dunn_ceramic_df, format = "markdown", caption = "Dunn Test Results for Ceramic Types")
  
  # Optional: Display interactive table
  datatable(
    dunn_ceramic_df,
    caption = "Interactive Dunn Test Results for Ceramic Types",
    options = list(pageLength = 5, autoWidth = TRUE, dom = 'Bfrtip', buttons = c('csv', 'excel')),
    extensions = 'Buttons'
  )
} else {
  cat("The Kruskal-Wallis test is not significant (p >= 0.05) for Ceramic Types.\n")
}
```

## 3.9.2. Non-parametric tests for the normalized data set by occupation period (mineral tempers)
```{r}
# Perform Kruskal-Wallis test for Tempers
kruskal_tempers <- combined_data %>%
  filter(Attribute == "Tempers") %>%
  na.omit() %>%
  kruskal.test(Normalized_Betweenness ~ Period, data = .)

# Print Kruskal-Wallis test results
cat("**Kruskal-Wallis Test Results for Tempers:**\n")
cat(sprintf("Chi-squared = %.3f, df = %d, p-value = %.5f\n\n",
            kruskal_tempers$statistic, kruskal_tempers$parameter, kruskal_tempers$p.value))

# Check if the test is significant
if (kruskal_tempers$p.value < 0.05) {
  cat("The Kruskal-Wallis test is significant (p < 0.05). Performing Dunn post hoc test...\n\n")
  
  # Perform Dunn post hoc test
  dunn_tempers <- combined_data %>%
    filter(Attribute == "Tempers") %>%
    na.omit() %>%
    dunnTest(Normalized_Betweenness ~ Period, data = ., method = "bonferroni")
  
  # Convert results to data frame for display
  dunn_tempers_df <- as.data.frame(dunn_tempers$res)
  
  # Print the Dunn test results in R Markdown
  cat("**Dunn Test Results for Tempers:**\n")
  knitr::kable(dunn_tempers_df, format = "markdown", caption = "Dunn Test Results for Tempers")
  
  # Optional: Display interactive table
  datatable(
    dunn_tempers_df,
    caption = "Interactive Dunn Test Results for Tempers",
    options = list(pageLength = 5, autoWidth = TRUE, dom = 'Bfrtip', buttons = c('csv', 'excel')),
    extensions = 'Buttons'
  )
} else {
  cat("The Kruskal-Wallis test is not significant (p >= 0.05) for Tempers.\n")
}
```

# 4. Spatial patterns of betweenness centrality by occupation period
## 4.1. Figure 7 (Measures of unusualness by Z-scores)
```{r}
# Set paths
shapefile_files <- c(
  "POLYGON_SURVEY_UNTILL_2023.shp",
  "POLYGON_SURVEY_UNTILL_2023.dbf",
  "POLYGON_SURVEY_UNTILL_2023.shx",
  "POLYGON_SURVEY_UNTILL_2023.prj",
  "POLYGON_SURVEY_UNTILL_2023.cpg"
)

# Base URL for the "GIS" folder in the repository "Network_Analysis_in_Tairona_Chiefdoms"
gis_base_url <- "https://raw.githubusercontent.com/luismiguelsoto/Network_Analysis_in_Tairona_Chiefdoms/main/GIS/"

# Temporary directory for shapefile components
temp_shp_dir <- tempdir()

for (f in shapefile_files) {
  shp_url  <- paste0(gis_base_url, utils::URLencode(f, reserved = TRUE))
  destfile <- file.path(temp_shp_dir, f)
  cat("Downloading shapefile component:", shp_url, "\n")
  download.file(shp_url, destfile = destfile, mode = "wb")
}

# Path to the main .shp file
shp_path <- file.path(temp_shp_dir, "POLYGON_SURVEY_UNTILL_2023.shp")
study_polygon <- st_read(shp_path)
print(study_polygon)

# Binarize similarity matrix
binarize_matrix <- function(matrix, thresh = 0.75) {
  matrix_binary <- matrix
  matrix_binary[matrix >= thresh] <- 1
  matrix_binary[matrix < thresh] <- 0
  diag(matrix_binary) <- 0
  return(matrix_binary)
}

# Calculate Z-score limits
calculate_z_limits <- function(z_scores) {
  min_z <- floor(min(z_scores, na.rm = TRUE))
  max_z <- ceiling(max(z_scores, na.rm = TRUE))
  return(c(min_z, max_z))
}

# Main plotting function
analyze_and_plot_spatial_network <- function(data, attr_data, title, coord) {
  # Calculate the similarity matrix and get valid row names
  BR_matrix <- sim.mat(data)
  valid_rows <- rownames(BR_matrix)
  
  # Subset attr_data and coord to match the valid vertices
  attr_data_sub <- attr_data[match(valid_rows, rownames(attr_data)), ]
  coord_sub <- coord[match(valid_rows, rownames(attr_data)), ]
  
  # Binarize the matrix and build the network
  BR_matrix_binary <- binarize_matrix(BR_matrix, thresh = 0.75)
  BR_net <- network(BR_matrix_binary, directed = FALSE)
  bw <- sna::betweenness(BR_net, gmode = "graph")
  
  mean_bw <- mean(bw, na.rm = TRUE)
  sd_bw <- sd(bw, na.rm = TRUE)
  z_scores <- (bw - mean_bw) / sd_bw
  z_limits <- calculate_z_limits(z_scores)
  
  # Ensure Site_Type has the desired order
  attr_data_sub$Site_Type <- factor(attr_data_sub$Site_Type, levels = c("Large sites", "Medium sites", "Small sites"))
  
  # Convert network to tidygraph and assign attributes
  BR_net_tidy <- as_tbl_graph(BR_net) %>%
    activate(nodes) %>%
    mutate(
      z_score = z_scores,
      x = coord_sub[, 1],
      y = coord_sub[, 2],
      Site_Type = attr_data_sub$Site_Type
    ) %>%
    filter(!is.na(x) & !is.na(y))
  
  # Build edges data frame (opcional, no se usará en el plot)
  edges_df <- as.data.frame(as_tbl_graph(BR_net) %>% activate(edges)) %>%
    mutate(
      from = as.character(from),
      to = as.character(to)
    ) %>%
    left_join(as.data.frame(BR_net_tidy) %>% select(name, x, y), by = c("from" = "name")) %>%
    rename(x = x, y = y) %>%
    left_join(as.data.frame(BR_net_tidy) %>% select(name, x, y), by = c("to" = "name"), suffix = c("", ".end")) %>%
    rename(xend = x.end, yend = y.end) %>%
    filter(!is.na(x) & !is.na(y) & !is.na(xend) & !is.na(yend))
  
  # Create the plot without geom_segment (only points)
  p <- ggplot() +
    geom_sf(data = study_polygon, fill = "gray90", color = "black", alpha = 0.5) +
    geom_point(data = as.data.frame(BR_net_tidy), 
               aes(x = x, y = y, size = abs(z_score), color = Site_Type)) +
    scale_color_manual(
      values = c("Large sites" = "red", "Medium sites" = "green", "Small sites" = "blue"),
      name = "Site Type"
    ) +
    scale_size_continuous(
      name = "Z-score\nBetweenness",
      limits = z_limits,
      breaks = seq(z_limits[1], z_limits[2], length.out = 5),
      labels = round(seq(z_limits[1], z_limits[2], length.out = 5), 2)
    ) +
    ggspatial::annotation_scale(location = "bl", width_hint = 0.5) +
    ggspatial::annotation_north_arrow(
      location = "tr", which_north = "true", 
      style = north_arrow_fancy_orienteering()
    ) +
    labs(title = title, x = "Longitude", y = "Latitude") +
    theme_minimal()
  
  return(p)
}
```

## 4.2.1. Neguanje: Ceramic Types
```{r}
Neguanje_Tipos <- as.data.frame(Neguanje_Tipos)
Neguanje_Tipos_attr <- as.data.frame(Neguanje_Tipos_attr)
rownames(Neguanje_Tipos) <- as.character(1:nrow(Neguanje_Tipos))
rownames(Neguanje_Tipos_attr) <- rownames(Neguanje_Tipos)
#
plot_neguanje_ceramic <- analyze_and_plot_spatial_network(
  Neguanje_Tipos,
  Neguanje_Tipos_attr,
  title = "Neguanje | Ceramic Types",
  coord = Neguanje_Tipos_attr[, 6:7]
)
print(plot_neguanje_ceramic)
```

## 4.2.2. Neguanje: Mineral Tempers
```{r}
Neguanje_Desgrasantes <- as.data.frame(Neguanje_Desgrasantes)
Neguanje_Desgrasantes_attr <- as.data.frame(Neguanje_Desgrasantes_attr)
rownames(Neguanje_Desgrasantes) <- as.character(1:nrow(Neguanje_Desgrasantes))
rownames(Neguanje_Desgrasantes_attr) <- rownames(Neguanje_Desgrasantes)
#
plot_neguanje_ceramic <- analyze_and_plot_spatial_network(
  Neguanje_Desgrasantes,
  Neguanje_Desgrasantes_attr,
  title = "Neguanje | Ceramic Types",
  coord = Neguanje_Desgrasantes_attr[, 10:11]
)
print(plot_neguanje_ceramic)
```

## 4.2.3. Buritaca: Ceramic Types
```{r}
Buritaca_Tipos <- as.data.frame(Buritaca_Tipos)
Buritaca_Tipos_attr <- as.data.frame(Buritaca_Tipos_attr)
rownames(Buritaca_Tipos) <- as.character(1:nrow(Buritaca_Tipos))
rownames(Buritaca_Tipos_attr) <- rownames(Buritaca_Tipos)
#
plot_Buritaca_ceramic <- analyze_and_plot_spatial_network(
  Buritaca_Tipos,
  Buritaca_Tipos_attr,
  title = "Buritaca | Ceramic Types",
  coord = Buritaca_Tipos_attr[, 5:6]
)
print(plot_Buritaca_ceramic)
```

## 4.2.4. Buritaca: Mineral Tempers
```{r}
Buritaca_Desgrasantes <- as.data.frame(Buritaca_Desgrasantes)
Buritaca_Desgrasantes_attr <- as.data.frame(Buritaca_Desgrasantes_attr)
rownames(Buritaca_Desgrasantes) <- as.character(1:nrow(Buritaca_Desgrasantes))
rownames(Buritaca_Desgrasantes_attr) <- rownames(Buritaca_Desgrasantes)
#
plot_Buritaca_tempers <- analyze_and_plot_spatial_network(
  Buritaca_Desgrasantes,
  Buritaca_Desgrasantes_attr,
  title = "Buritaca | Tempers",
  coord = Buritaca_Desgrasantes_attr[, 9:10]
)
print(plot_Buritaca_tempers)
```

## 4.2.5. Tairona: Ceramic Types
```{r}
Tairona_Tipos <- as.data.frame(Tairona_Tipos)
Tairona_Tipos_attr <- as.data.frame(Tairona_Tipos_attr)
rownames(Tairona_Tipos) <- as.character(1:nrow(Tairona_Tipos))
rownames(Tairona_Tipos_attr) <- rownames(Tairona_Tipos)
#
plot_Tairona_ceramic <- analyze_and_plot_spatial_network(
  Tairona_Tipos,
  Tairona_Tipos_attr,
  title = "Tairona | Ceramic Types",
  coord = Tairona_Tipos_attr[, 8:9]
)
print(plot_Tairona_ceramic)
```

## 4.2.4. Tairona: Mineral Tempers
```{r}
Tairona_Desgrasantes <- as.data.frame(Tairona_Desgrasantes)
Tairona_Desgrasantes_attr <- as.data.frame(Tairona_Desgrasantes_attr)
rownames(Tairona_Desgrasantes) <- as.character(1:nrow(Tairona_Desgrasantes))
rownames(Tairona_Desgrasantes_attr) <- rownames(Tairona_Desgrasantes)
#
plot_Tairona_tempers <- analyze_and_plot_spatial_network(
  Tairona_Desgrasantes,
  Tairona_Desgrasantes_attr,
  title = "Tairona | Tempers",
  coord = Tairona_Desgrasantes_attr[, 10:11]
)
print(plot_Tairona_tempers)
```
